---
title: "Review of R objects and operators"
date: "9/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This note is heavily borrowed from [Prof. Cosma Shalizi's course at CMU](http://www.stat.cmu.edu/~cshalizi/statcomp/14/)



## Different kinds of data object

All data is represented in binary format, by **bits** (TRUE/FALSE, YES/NO, 1/0)

- **Booleans** Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero), represented by a fixed-length block of bits
- **Characters** fixed-length blocks of bits, with special coding;
**strings** = sequences of characters
- **Floating point numbers**: a fraction (with a finite number of bits) times an exponent, like $1.87 \times {10}^{6}$, but in binary form
- **Missing or ill-defined values**: `NA`, `NaN`, etc.

## Operators


- **Unary** `-` for arithmetic negation, `!` for Boolean
- **Binary** usual arithmetic operators, plus ones for modulo and integer division; take two numbers and give a number



```{r}
7+5
7-5
7*5
7^5
```


```{r}
7/5
7 %% 5
7 %/% 5
```



### Comparisons

These are also binary operators; they take two objects, like numbers, and give a Boolean
```{r}
7 > 5
7 < 5
7 >= 7
7 <= 5
```


```{r}
7 == 5
7 != 5
```

**Boolean operators**

Basically "and" and "or":

```{r}
(5 > 7) & (6*7 == 42)
(5 > 7) | (6*7 == 42)
```

<small>(will see special doubled forms, `&&` and `||`, later)</small>


**More types**

`typeof()` function returns the type

`is.`_foo_`()` functions return Booleans for whether the argument is of type _foo_

`as.`_foo_`()` (tries to) "cast" its argument to type _foo_ --- to translate it sensibly into a _foo_-type value


```{r}
typeof(7)
is.numeric(7)
is.na(7)
is.na(7/0)
is.na(0/0)
```
<small>Why is 7/0 not NA, but 0/0 is?</small>

```{r}
is.character(7)
is.character("7")
is.character("seven")
is.na("seven")
```


```{r}
as.character(5/6)
as.numeric(as.character(5/6))
6*as.numeric(as.character(5/6))
5/6 == as.numeric(as.character(5/6))
```
<small>(why is that last FALSE?)</small>


### Assigning variables


We can give names to data objects; these give us **variables**

A few variables are built in:
```{r}
pi
```

Variables can be arguments to functions or operators, just like constants:
```{r}
pi*10
cos(pi)
```



Most variables are created with the **assignment operator**, `<-` or `=`  
```{r}
approx.pi <- 22/7
approx.pi
diameter.in.cubits = 10
approx.pi*diameter.in.cubits
```


The assignment operator also changes values:
```{r}
circumference.in.cubits <- approx.pi*diameter.in.cubits
circumference.in.cubits
circumference.in.cubits <- 30
circumference.in.cubits
```



Using names and variables makes code: easier to design, easier to debug, less prone to bugs, easier to improve,  and easier for others to read

Avoid "magic constants"; use named variables
<small>you will be graded on this!</small>

Named variables are a first step towards **abstraction**


### The workspace

What names have you defined values for?
```{r}
ls()
objects()
```

Getting of variables:
```{r}
rm("circumference.in.cubits")
ls()
```

### First data structure: vectors


Group related data values into one object, a **data structure**

A **vector** is a sequence of values, all of the same type
```{r}
x <- c(7, 8, 10, 45)
x
is.vector(x)
```

`c()` function returns a vector containing all its arguments in order

`x[1]` is the first element, `x[4]` is the 4th element  
`x[-4]` is a vector containing all but the fourth element


`vector(length=6)` returns an empty vector of length 6; helpful for filling things up later
```{r}
weekly.hours <- vector(length=5)
weekly.hours[5] <- 8
```

### Vector arithmetic


Operators apply to vectors "pairwise" or "elementwise":
```{r}
y <- c(-7, -8, -10, -45)
x+y
x*y
```

### Recycling

**Recycling** repeat elements in shorter vector when combined with longer
```{r}
x + c(-7,-8)
x^c(1,0,-1,0.5)
```

Single numbers are vectors of length 1 for purposes of recycling:
```{r}
2*x
```


Can also do pairwise comparisons:
```{r}
x > 9
```
Note: returns Boolean vector

Boolean operators work elementwise:
```{r}
(x > 9) & (x < 20)
```


To compare whole vectors, best to use `identical()` or `all.equal()`:
```{r}
x == -y
identical(x,-y)
identical(c(0.5-0.3,0.3-0.1),c(0.3-0.1,0.5-0.3))
all.equal(c(0.5-0.3,0.3-0.1),c(0.3-0.1,0.5-0.3))
```

### Functions on vectors


Lots of functions take vectors as arguments:
- `mean()`, `median()`, `sd()`, `var()`, `max()`, `min()`, `length()`, `sum()`: return single numbers
- `sort()` returns a new vector
- `hist()` takes a vector of numbers and produces a histogram, a highly structured object, with the side-effect of making a plot
- Similarly `ecdf()` produces a cumulative-density-function object
- `summary()` gives a five-number summary of numerical vectors
- `any()` and `all()` are useful on Boolean vectors

### Addressing vectors


Vector of indices:
```{r}
x[c(2,4)]
```

Vector of negative indices
```{r}
x[c(-1,-3)]
```
<small>(why that, and not  `8 10`?)</small>


Boolean vector:
```{r}
x[x>9]
y[x>9]
```

`which()` turns a Boolean vector in vector of TRUE indices:
```{r}
places <- which(x > 9)
places
y[places]
```

Named components


You can give names to elements or components of vectors
```{r}
names(x) <- c("v1","v2","v3","fred")
names(x)
x[c("fred","v1")]
```
note the labels in what R prints; not actually part of the value


`names(x)` is just another vector (of characters):
```{r}
names(y) <- names(x)
sort(names(x))
which(names(x)=="fred")
```




### Take-Aways

- We write programs by composing functions to manipulate data
- The basic data types let us represent Booleans, numbers, and characters
- Data structure let us group related values together
- Vectors let us group values of the same type
- Use variables rather a profusion of magic constants
- Name components of structures to make data more meaningful



### Peculiarites of floating-point numbers

The more bits in the fraction part, the more precision

The R floating-point data type is a `double`, a.k.a. `numeric`
<small>back when memory was expensive, the now-standard
number of bits was twice the default</small>

Finite precision $\Rightarrow$ arithmetic on `doubles` $\neq$ arithmetic on $\mathbb{R}$.


```{r}
0.45 == 3*0.15
0.45 - 3*0.15
```


### Often ignorable, but not always

- Rounding errors tend to accumulate in long calculations
- When results should be $\approx 0$, errors can flip signs
- Usually better to use `all.equal()` than exact comparison
```{r}
(0.5 - 0.3) == (0.3 - 0.1)
all.equal(0.5-0.3, 0.3-0.1)
```

### Peculiarities of Integers


Typing a whole number in the terminal doesn't make an integer; it makes a double, whose fractional part is 0
```{r}
is.integer(7)
```

This looks like an integer
```{r}
as.integer(7)
```

To test for being a whole number, use `round()`:
```{r}
round(7) == 7
```